Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-13 paper/main.tex
minor

One might think that our set of potential functions is too simplistic
to be able to express and prove bounds for realistic
programs. Nevertheless, we can handle challenging example programs
without special tricks or techniques.  Examples \emph{speed\_1} and
\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
demonstrate that our method can handle \emph{tricky iteration
  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
bounds as our analysis but requires heuristics for its counter
instrumentation.  These loops can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
generate.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

@@

Another simple example is given on \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are in blue in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $lg >
\log_2(h-l)$ as an invariant.  This technique allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$.

@@

Example \emph{t13} shows how amortization can be used to find linear
bounds for nested loops.  The outer loop is iterated $|[0,x]|$ times.
In the conditional, we either (the branching condition is arbitrary)
increment the variable $y$ or we execute an inner loop in which $y$ is
counted back to $0$.  $\toolname$ computes a tight bound.
\iffull{Again, the constants $0$ and $1$ in the
  inner loop can as well be replace by something more interesting, say
  $9$ and $10$ like in example \emph{t08}.  Then we still obtain a
  tight linear bound.}{}
%
\ifdefined\fullversion
\begin{figure}[t]
  \centering
    \begin{minipage}[b]{.9\linewidth}
    \begin{center}
   \begin{lstlisting}
int srch(
  int t[], int n,  /* haystack */
  int p[], int m,  /* needle */
  int b[]
) { int i=0, j=0, k=-1;
    while (i < n) {
      while (j >= 0 && t[i]!=p[j]) {
        k = b[j];
        assert(k > 0 && k <= j + 1);
        j -= k; tick(1)
      }
      i++, j++;
      if (j == m) break;
      tick(1);
    }
    return i;
  }
   \end{lstlisting}
\vspace{-1.2cm}
\hspace{3.8cm}
\footnotesize
\begin{tabular}{|rl|}
\hline
\toolname & $1 {+} 2|\inter 0 n|$ \\
Rank & $O(n^2)$ \\
LOOPUS & $2 {+} 3\max(n, 0)$\\
\hline
\end{tabular}
    \end{center}
\vspace{-0.1cm}
  \end{minipage}
  \caption{The Knuth-Morris-Pratt algorithm for string search.}
  \label{fig:KMP}
\end{figure}

@@

Since we have a formal cost semantics, we can run our examples with
this semantics for different inputs and measure the cost to compare it
to our derived bound.  \pref{fig:3d} shows such a comparison for
example \emph{t08}, a variant of \emph{t08a} from \pref{sec:AAA}.  One
can see that the derived constant factors are the best possible if the
input variable $x$ is non-negative.  \iffull{If $x$ is negative then
  the bound is only slightly off.}{}

@@

\begin{figure}[t]
\center
\vspace{-.2cm}
\includegraphics[width=.9\linewidth]{fig/bound3d}
\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
  (blue lines) and the measured runtime cost (red crosses) for example
  \emph{t08}. For $x\ge 0$ the bound is tight.}
\label{fig:3d}
\end{figure}

@@

In example \emph{t10} we also do not restrict the inputs $x$ and $y$.
They can be negative, positive, or zero.  The star {\tt *} in the
conditional, stands for an arbitrary assertion.  In each branch of the
conditional we can obtain the constant potential $1$ since the interval
size $|[y,x]|$ is decreasing.

Example \emph{t13} shows how amortization can be used to handle tricky
nested loops.  The outer loop is iterated $|[0,x]|$ times.  In the
conditional, we either (the branching condition is again arbitrary)
increment the variable $y$ or we execute an inner loop in which $y$ is
counted back to $0$.  The analysis computes a tight linear bound for
this program.  Again, the constants $0$ and $1$ in the inner loop can
as well be replace by something more interesting, say $9$ and $10$
like in example \emph{t08}.  Then we still obtain a tight linear
bound.

Example \emph{t27} is similar to example \emph{t13}.  Instead of
decrementing the variable $x$ in the outer loop we this time increment
the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
we increment the variable $y$ by $1000$.  We then execute an inner
loop that increments $y$ by $100$ until $y=0$.  The analysis can
derive that only the first execution of the inner loop depends on the
initial value of $y$.  We again derive a tight bound.

@@

Example \emph{t16} is an extension of example \emph{t15}. We again
assume that $y$ is non-negative and use the same mechanism to iterate
the outer loop as in \emph{t15}.  In the inner loop, we also count the
variable $z$ down to zero and perform $|[0,z]|$ iterations.  However,
instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
computes the linear bound $101|[0,x]|$.  The
assignment of potential to the size interval $|[0,x]|$ in instead of
$|[y,x]|$ is a random choice of the LP solver.

@@

At first sight, example \emph{t30} appears to be a simple loop that
decrements the variable $x$ down to zero.  However, a closer look
reveals that the loop actually decrements both input variables $x$ and
$y$ down to zero before terminating.  In the loop body, first $x$ is
decremented by one.  Then the values of the variables $x$ and $y$ are
switched using the local variable $t$ as a buffer.  Our analysis
infers the tight bound $|[0,x]|+|[0,y]|$.

@@

Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
al~\cite{GulwaniMC09}.  They are both handled by the SPEED tool but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{fig2\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{fig2\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

Example \emph{nested\_multiple} is similar to example \emph{fig2\_1}.
Instead of incrementing variable $y$ in the outer loop, $y$ is here
potentially incremented multiple times in each iteration of the outer
loop.  The idea of example \emph{nested\_single} is similar.  However,
instead of incrementing variable $y$ in the inner loop, we increment
$x$, the counter variable of the outer loop. Our analyzer derives a
tight bound for both programs.  Note that a star \emph{*} in a
branching condition denotes an arbitrary boolean condition that might
of course change while iterating (non-deterministic choice).

Example \emph{sequential\_single} is like example
\emph{nested\_single}.  The only difference is that the inner loop of
\emph{nested\_single} is now evaluated after the outer loop.  Example
\emph{simple\_multiple} is a variant of example \emph{fig2\_1} and
\emph{simple\_single} is a simple variant of \emph{nested\_single}.
We derive tight bounds for all aforementioned programs.

@@

In example \emph{fig4\_4} it is essential that $m$ is positive.  That
ensures that we can obtain constant potential for the interval size
$|[0,i]|$ in the \emph{else} branch of the conditional since $|[0,i]|$
decreases.  Example \emph{fig4\_5} is an examples that we can not
handle automatically.  The execution is bounded because the boolean
value of the test \code{dir==fwd} does not change during the iteration
of the loop.  As a result, the variable $i$ is either counted down to
$0$ or up to $n$.  Our tool cannot handle example \emph{fig4\_5}
because we don't do path sensitive reasoning.  Note however that it
would be more efficient to move the test \code{dir==fwd} outside of
the loop (this would be also done by an optimizing compiler).  The
resulting program can then be analyzed by our tool.

@@

Similar to example \emph{ex1}, example \emph{ex2}
form~\cite{GulwaniZ10} focuses on the number of iterations of the
\emph{outer loop}.  While the whole program is not terminating, the
number of iterations of the outer loop is bounded by $|[0,n]|$.
Finally, example \emph{ex2} is similar to example
\emph{nested\_single}, and \emph{ex3} is a variant of example
\emph{t47}.

@@

Example \emph{t39} uses mutual recursion.  The function
\code{count\_down} is similar to the function with the same name in
example \emph{t37}.  However, we do not count down to $0$ but to a
variable $y$ that is passed as an argument and we call the function
\code{count\_up} afterwards.  The function \code{count\_up} is dual to
\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
\code{count\_down}.  For the function \code{main}, which calls
\code{count\_down(y,z)}, the analyzer computes the tight bound $1.33 +
0.67 |[z,y]|$.

@@

Example \emph{Greatest Common Divisor} is the usual implementation of
the GCD algorithm.  We automatically derive a linear bound.  Note that
the two tests at the beginning that check if the inputs are positive
are essential.  Finally, example \emph{Quick Sort} shows a skeleton of
the quick sort sorting algorithm.  Since we can only derive linear
bounds we left out the inner loop that swaps the array elements and
determines the position \code{n+lo} of the pivot.  We only assert that
\code{lo < n+lo <= hi}.  We then derive a tight linear bound.

>>>>

One might think that our set of potential functions is too simplistic
to be able to express and prove bounds for realistic
programs. Nevertheless, we can handle challenging example programs
without special tricks or techniques.  Examples \emph{speed\_1} and
\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
demonstrate that our method can handle \emph{tricky iteration
  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
bounds as our analysis but requires heuristics for its counter
instrumentation.  These loops can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
generate.
%
In Example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

@@

Another simple example is given in \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are in blue in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $lg >
\log_2(h-l)$ as an invariant.  This technique allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$.

@@

Example \emph{t13} shows how amortization can be used to find linear
bounds for nested loops.  The outer loop is iterated $|[0,x]|$ times.
In the conditional, we either (the branching condition is arbitrary)
increment the variable $y$ or we execute an inner loop in which $y$ is
counted back to $0$.  $\toolname$ computes a tight bound.
\iffull{Again, the constants $0$ and $1$ in the
  inner loop can as well be replace by something more interesting, say
  $9$ and $10$ like in Example \emph{t08}.  Then we still obtain a
  tight linear bound.}{}
%
\ifdefined\fullversion
\begin{figure}[t]
  \centering
    \begin{minipage}[b]{.9\linewidth}
    \begin{center}
   \begin{lstlisting}
int srch(
  int t[], int n,  /* haystack */
  int p[], int m,  /* needle */
  int b[]
) { int i=0, j=0, k=-1;
    while (i < n) {
      while (j >= 0 && t[i]!=p[j]) {
        k = b[j];
        assert(k > 0 && k <= j + 1);
        j -= k; tick(1)
      }
      i++, j++;
      if (j == m) break;
      tick(1);
    }
    return i;
  }
   \end{lstlisting}
\vspace{-1.2cm}
\hspace{3.8cm}
\footnotesize
\begin{tabular}{|rl|}
\hline
\toolname & $1 {+} 2|\inter 0 n|$ \\
Rank & $O(n^2)$ \\
LOOPUS & $2 {+} 3\max(n, 0)$\\
\hline
\end{tabular}
    \end{center}
\vspace{-0.1cm}
  \end{minipage}
  \caption{The Knuth-Morris-Pratt algorithm for string search.}
  \label{fig:KMP}
\end{figure}

@@

Since we have a formal cost semantics, we can run our examples with
this semantics for different inputs and measure the cost to compare it
to our derived bound.  \pref{fig:3d} shows such a comparison for
Example \emph{t08}, a variant of \emph{t08a} from \pref{sec:AAA}.  One
can see that the derived constant factors are the best possible if the
input variable $x$ is non-negative.  \iffull{If $x$ is negative then
  the bound is only slightly off.}{}

@@

\begin{figure}[t]
\center
\vspace{-.2cm}
\includegraphics[width=.9\linewidth]{fig/bound3d}
\caption{The automatically derived bound $1.33|[x,y]| + 0.33 |[0,x]|$
  (blue lines) and the measured runtime cost (red crosses) for Example
  \emph{t08}. For $x\ge 0$ the bound is tight.}
\label{fig:3d}
\end{figure}

@@

In Example \emph{t10} we also do not restrict the inputs $x$ and $y$.
They can be negative, positive, or zero.  The star {\tt *} in the
conditional, stands for an arbitrary assertion.  In each branch of the
conditional we can obtain the constant potential $1$ since the interval
size $|[y,x]|$ is decreasing.

Example \emph{t13} shows how amortization can be used to handle tricky
nested loops.  The outer loop is iterated $|[0,x]|$ times.  In the
conditional, we either (the branching condition is again arbitrary)
increment the variable $y$ or we execute an inner loop in which $y$ is
counted back to $0$.  The analysis computes a tight linear bound for
this program.  Again, the constants $0$ and $1$ in the inner loop can
as well be replace by something more interesting, say $9$ and $10$
like in Example \emph{t08}.  Then we still obtain a tight linear
bound.

Example \emph{t27} is similar to Example \emph{t13}.  Instead of
decrementing the variable $x$ in the outer loop we this time increment
the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
we increment the variable $y$ by $1000$.  We then execute an inner
loop that increments $y$ by $100$ until $y=0$.  The analysis can
derive that only the first execution of the inner loop depends on the
initial value of $y$.  We again derive a tight bound.

@@

Example \emph{t16} is an extension of Example \emph{t15}. We again
assume that $y$ is non-negative and use the same mechanism to iterate
the outer loop as in \emph{t15}.  In the inner loop, we also count the
variable $z$ down to zero and perform $|[0,z]|$ iterations.  However,
instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
computes the linear bound $101|[0,x]|$.  The
assignment of potential to the size interval $|[0,x]|$ in instead of
$|[y,x]|$ is a random choice of the LP solver.

@@

At first sight, Example \emph{t30} appears to be a simple loop that
decrements the variable $x$ down to zero.  However, a closer look
reveals that the loop actually decrements both input variables $x$ and
$y$ down to zero before terminating.  In the loop body, first $x$ is
decremented by one.  Then the values of the variables $x$ and $y$ are
switched using the local variable $t$ as a buffer.  Our analysis
infers the tight bound $|[0,x]|+|[0,y]|$.

@@

Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
al~\cite{GulwaniMC09}.  They are both handled by the SPEED tool but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In Example \emph{fig2\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{fig2\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

Example \emph{nested\_multiple} is similar to Example \emph{fig2\_1}.
Instead of incrementing variable $y$ in the outer loop, $y$ is here
potentially incremented multiple times in each iteration of the outer
loop.  The idea of Example \emph{nested\_single} is similar.  However,
instead of incrementing variable $y$ in the inner loop, we increment
$x$, the counter variable of the outer loop. Our analyzer derives a
tight bound for both programs.  Note that a star \emph{*} in a
branching condition denotes an arbitrary boolean condition that might
of course change while iterating (non-deterministic choice).

Example \emph{sequential\_single} is like Example
\emph{nested\_single}.  The only difference is that the inner loop of
\emph{nested\_single} is now evaluated after the outer loop.  Example
\emph{simple\_multiple} is a variant of Example \emph{fig2\_1} and
\emph{simple\_single} is a simple variant of \emph{nested\_single}.
We derive tight bounds for all aforementioned programs.

@@

In Example \emph{fig4\_4} it is essential that $m$ is positive.  That
ensures that we can obtain constant potential for the interval size
$|[0,i]|$ in the \emph{else} branch of the conditional since $|[0,i]|$
decreases.  Example \emph{fig4\_5} is an examples that we can not
handle automatically.  The execution is bounded because the boolean
value of the test \code{dir==fwd} does not change during the iteration
of the loop.  As a result, the variable $i$ is either counted down to
$0$ or up to $n$.  Our tool cannot handle Example \emph{fig4\_5}
because we don't do path sensitive reasoning.  Note however that it
would be more efficient to move the test \code{dir==fwd} outside of
the loop (this would be also done by an optimizing compiler).  The
resulting program can then be analyzed by our tool.

@@

Similar to Example \emph{ex1}, Example \emph{ex2}
form~\cite{GulwaniZ10} focuses on the number of iterations of the
\emph{outer loop}.  While the whole program is not terminating, the
number of iterations of the outer loop is bounded by $|[0,n]|$.
Finally, Example \emph{ex2} is similar to example
\emph{nested\_single}, and \emph{ex3} is a variant of Example
\emph{t47}.

@@

Example \emph{t39} uses mutual recursion.  The function
\code{count\_down} is similar to the function with the same name in
Example \emph{t37}.  However, we do not count down to $0$ but to a
variable $y$ that is passed as an argument and we call the function
\code{count\_up} afterwards.  The function \code{count\_up} is dual to
\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
\code{count\_down}.  For the function \code{main}, which calls
\code{count\_down(y,z)}, the analyzer computes the tight bound $1.33 +
0.67 |[z,y]|$.

@@

Example \emph{Greatest Common Divisor} is the usual implementation of
the GCD algorithm.  We automatically derive a linear bound.  Note that
the two tests at the beginning that check if the inputs are positive
are essential.  Finally, Example \emph{Quick Sort} shows a skeleton of
the quick sort sorting algorithm.  Since we can only derive linear
bounds we left out the inner loop that swaps the array elements and
determines the position \code{n+lo} of the pivot.  We only assert that
\code{lo < n+lo <= hi}.  We then derive a tight linear bound.
